---
title: "Exploratory Analysis SAT scores in Public Highschools"
output:
  html_document:
    fig_height: 4
    fig_width: 7
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Reading the data
Load data
```{r}
data <- read.csv("MA_Public_Schools_2017.csv")
str(data)
```
## factors are present
```{r}
summary(data$AP_Test.Takers)
summary(data$Grade)  
```

## Subsetting to Highschools and variables of numerical value
```{r}
hs_data <- subset(data, X12_Enrollment > 0)
num <- sapply(hs_data, is.numeric)
hs_corr_data <- hs_data[,num]

hs_corr_data$AP_Test.Takers <- as.numeric(hs_data$AP_Test.Takers)
hs_corr_data$AP_Tests.Taken <- as.numeric(hs_data$AP_Tests.Taken)
hs_corr_data_cl <- hs_corr_data[colSums(!is.na(hs_corr_data)) > 40]
```

## Selecting numerical values from data for analysis
```{r}
hs_data_ready<- 
  select(hs_corr_data_cl, 
         #School.Code, Zip, 
         #District.Code,
         #SP_Enrollment, 
         TOTAL_Enrollment, 
         X..First.Language.Not.English,
         X..High.Needs,
         X..Economically.Disadvantaged,
         X..African.American,
         X..Asian,
         X..Hispanic,
         X..White,
         X..Multi.Race..Non.Hispanic,
         X..Males,
         X..Females,
         Average.Class.Size,
         Average.Salary,
         Average.In.District.Expenditures.per.Pupil,
         Average.Expenditures.per.Pupil,
         X..in.Cohort,
         X..GED,
         X..Non.Grad.Completers,
         X..Dropped.Out,
         X..Attending.College,
         X..Private.Four.Year,
         X..Public.Four.Year,
         X..MA.Community.College,
         X..AP_Score.3.5,
         SAT_Tests.Taken,
         Average.SAT_Reading,
         Average.SAT_Math) 
```

## Exploring Missing Data

Public schools with missing SAT & AP scores tend to come from schools with higher % of Economically disadvantaged students

The data-set only contains 394 high-schools, and the total number of high schools missing SAT test scores is 62. Meaning if we simply drop the schools with missing SAT scores I lose about 16% of my total observations.

```{r}

Data_na_scores <-
  hs_data_ready %>% 
  select( everything() ) %>%
  filter(is.na(Average.SAT_Math) 
         & is.na(Average.SAT_Reading) 
         & is.na(X..AP_Score.3.5))
```

### All Schools vs. Schools with missing scores
```{r}

summary(hs_data_ready$X..Economically.Disadvantaged)


hs_data_ready %>%
  ggplot(aes(x=X..Economically.Disadvantaged)) + 
  ggtitle("All Schools") +
  geom_histogram(bins=20, fill = "red") +
  theme_bw()+theme(axis.title = element_text(size=16),axis.text = element_text(size=14))+
  ylab("Count") 


Data_na_scores %>%
  ggplot(aes(x=X..Economically.Disadvantaged)) + 
  ggtitle("Schools with missing scores") +
  geom_histogram(bins=20, fill = "red") +
  theme_bw()+theme(axis.title = element_text(size=16),axis.text = element_text(size=14))+
  ylab("Count") 
```

Instead I used the K Nearest Neighbor algorithm to impute the missing values. 


#let us normalize & impute

```{r}
library(caret)


trans_impute <- preProcess(hs_data_ready,
                           method = c("BoxCox", "center", "scale", "knnImpute"))
transformed_imputed <- predict(trans_impute, hs_data_ready)
```


### Comparing Original data and imputed & normalized data

Original Data
``` {r}

boxplot(hs_data_ready$X..Economically.Disadvantaged)

hs_data_ready %>%
  ggplot(aes(x=X..Economically.Disadvantaged)) + 
  ggtitle("Original Data") +
  geom_histogram(bins=20, fill = "red") +
  theme_bw()+theme(axis.title = element_text(size=16),axis.text = element_text(size=14))+
  ylab("Count") 
```

Imputed and Normalized data
``` {r}
boxplot(transformed_imputed$X..Economically.Disadvantaged)

transformed_imputed %>%
  ggplot(aes(x=X..Economically.Disadvantaged)) + 
  ggtitle("Imputed & Normalized data") +
  geom_histogram(bins=20, fill = "red") +
  theme_bw()+theme(axis.title = element_text(size=16),axis.text = element_text(size=14))+
  ylab("Count") 
```

## Relationships of Interest 

SAT Math scores were the variable of interest in these examples, since AP scores, SAT Reading and SAT Math are all highly correlated, it doesn’t really matter which of the 3 we use

```{r}
hs_data_quick<- 
  select(transformed_imputed, 
         TOTAL_Enrollment,
         X..High.Needs,
         X..Economically.Disadvantaged,
         Average.Class.Size,
         Average.Salary,
         Average.In.District.Expenditures.per.Pupil,
         Average.Expenditures.per.Pupil,
         Average.SAT_Math)

correlations_quick<-cor(x=hs_data_quick, use = "pairwise")
corrplot(correlations_quick, method = "circle", order= "hclust")
```

Expenditures per pupil are highly correlated with teacher salary, but not correlated at all with class size. Seems like teacher salary affects the public budget a lot more than class size does.

Average SAT math shows a slight negative correlation with expenditure variables. This may be due to higher expenditures in schools with more “high needs” students (as shown by correlation above), thus the expenditure vs. SAT score relationship is really dominated by the socioeconomic vs. SAT score relationship ( I will come back to this)

```{r}
a<- ggplot(transformed_imputed, aes(Average.SAT_Math, X..Economically.Disadvantaged))
a + geom_point()+
  geom_smooth(method = "lm", se = FALSE)+
  labs(title = "% of Economically Disadvantaged vs. Average SAT Math Scores")

b<- ggplot(transformed_imputed, aes(Average.SAT_Math, Average.Salary))
b + geom_point() + 
  geom_smooth(method = "lm", se = FALSE)+
  labs(title = "Average.Salary vs. Average SAT Math Scores")

c<- ggplot(transformed_imputed, aes(Average.SAT_Math, Average.Class.Size))
c + geom_point()+ 
  geom_smooth(method = "lm", se = FALSE)+
  labs(title = "Average.Class.Size vs. Average SAT Math Scores")

d<- ggplot(transformed_imputed, aes(Average.SAT_Math, Average.Expenditures.per.Pupil))
d + geom_point()+ 
  geom_smooth(method = "lm", se = FALSE)+
  labs(title = "Average.Expenditures.per.Pupil vs. Average SAT Math Scores")

e<- ggplot(transformed_imputed, aes(Average.SAT_Math, TOTAL_Enrollment))
e + geom_point() + geom_smooth(method = "lm", se = FALSE)+
  labs(title = "Total Enrollment vs. Average SAT Math Scores")
  
```
Scatter plots show a stronger relationship between Average SAT scores, (Expenditures per pupil & Total enrollment) than (Average Salary & Average Class size).

Let’s test these two groups while controlling for Economic disadvantage.

note: Since the two groups are correlated, I tested them by running two different regressions and seeing how much each variable contributed to SAT scores while controlling for Economic disadvantage.

linear_cluster1 <- lm(Average.SAT_Math~ X..Economically.Disadvantaged + Average.Salary + Average.Class.Size, data = hs_data_Cluster1)
summary(linear_cluster1)

